---
title: "Ball Helix Bioinformatics Assoc. Scientist Analyst Test"
author: "M. Ross Alexander"
date: "12/13/2019"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---
```{r libraries, echo=F, messages=F, warning=F}
# BiocManager::install("pasilla")
# BiocManager::install("DESeq")
# BiocManager::install("edgeR")
# BiocManager::install("GO.db")
# BiocManager::install("org.Hs.eg.db")

library(DESeq) # version 1.9.11
library(edgeR) # version 2.99.8
library(VennDiagram)
library(GO.db)
library(org.Hs.eg.db)
library(ggplot2) 

```

## Analyst Test data exercise
The original data frame consisted of count data produced by the tool Salmon for three replicates during the bud phase and then two days after flowering. The following document walks through the analyses used to determine which genes might be good candidates for further study. 

*basic folder structure*
figures--preliminary figures saved here
input_data--original data files to be called for analyses
processed_data--modified data files produced by analyses

## Summary
##### 1) This analysis identified 8245 genes that showed statistically significant changes between bud and flower phases (alpha = 0.05).  From these initial 8245 genes, I identified 99 genes that were not present in the bud phase but up regulated in the flower phase and 108 genes that were present in the bud phase but then down regulated in the flower phase. These 207 genes appear to be good candidates for further investigation of the desred trait

##### 2) The figure illustrating the position of these genes can found in *figures/gene_onoff.png*.

##### 3) For this experiment three samples were taken from plants in the bud phase and then two days after flowering. Each of these samples were analyzed to identify genes of interest using the tool Salmon. Using various references, this tool identifies gene sequences of interest and allows us to count how often they are expressed in our samples.  The higher the count, the more these genes are present in each the bud and the flowering phases. By tallying these counts we have identified genes that were either turned on or off as the plants went from the bud phase (closed flowers) to 2 days after the flowering phase. 

## Included Scripts

This summary is composed of insights cleaned after carrying out the following scripts. Not all analyses detailed in scripts are presented in this document.

1_data_walk through.R--Initial looks at the data frame. Included to show thought process; considered rough.

2_genetics_example.R--preliminary analyses to identify potential genes of interest.

Original data frame
```{r, message=F}
# Read in data ------------------------------------------------------------

datafile = read.table("input_data/test_gene_expression_matrix.txt", header=T, row.names = 1) # making column 1 row names
head(datafile)
summary(datafile)

## Read in the data making the row names the first column
counttable <- datafile #creating a count table for later use
head(counttable)
```
Making a metadata file to run edgeR analytical stemps
```{r, echo=FALSE, warning=FALSE, messages=F}
## Make metadata data.frame
meta <- data.frame(
  row.names=colnames(counttable),
  condition=c("bud", "bud", "bud", "flower", "flower", "flower"))
meta$condition <- relevel(meta$condition, ref="bud")
meta
```

```{r, echo=FALSE, warning=FALSE, fig.height=8, fig.width=8, messages=F}
## Make design matrix
condition <- relevel(factor(meta$condition), ref="bud")
#libType <- factor(meta$libType)
edesign <- model.matrix(~condition)

## Make new DGEList, normalize by library size, and estimate dispersion allowing possible trend with average count size
e <- DGEList(counts=counttable)
e <- calcNormFactors(e)
e <- estimateGLMCommonDisp(e, edesign)
e <- estimateGLMTrendedDisp(e, edesign) 
e <- estimateGLMTagwiseDisp(e, edesign)

## MDS Plot
plotMDS(e, main="edgeR MDS Plot")
```

#### The ordination plot shows us that the genes being expressed during the bud phase are more similar across replicates than the genes being expressed after flowering

```{r,echo=FALSE, fig.height=20, fig.width=15, messages=F}

## Biological coefficient of variation plot
# plotBCV(e, cex=0.4, main="edgeR: Biological coefficient of variation (BCV) vs abundance")

## Fit the model, testing the coefficient for the treated vs untreated comparison
efit <- glmFit(e, edesign)
efit <- glmLRT(efit, coef="conditionflower")

## Make a table of results
etable <- topTags(efit, n=nrow(e))$table
etable <- etable[order(etable$FDR), ]
head(etable)


# Looking at which factors were significant
addmargins(table(sig_edgeR=etable$FDR<0.05))
summary(decideTests(efit)) # about 4k down regulated; about 4k up regulated @ 0.05FDR
# PLot Log-fold change against log-counts per million with DE genes highlighted
plotMD(efit)
abline(h=c(-1,1), col="blue")
```

#### This plot shows the genes in which we see a significant count change that either increases (up; red) or decreases (down; blue).  Blue horizontal lines show a 2-fold change in expression. Significance was determined at the alpha = 0.05 level.



```{r,echo=FALSE, fig.height=20, fig.width=15, messages=F, warning=FALSE}
# Want to pull out the names of the genes that were up regulated and down regulated
test <- as.data.frame(decideTests(efit))
test$gene <- as.factor(row.names(test))
summary(test)
up.reg <- test[test$conditionflower > 0,]
down.reg <- test[test$conditionflower < 0,]

summary(up.reg)
summary(down.reg)

```

#### Taking the underlying data frame that made the previous figure, we can extract the names of genes that were significantly up or down regulated.  This still leaves about four thousand genes each that showed significant increases or decreases in expression between buds and flowers.  Although fewer than the total genes we started with (ca. 30,000), we can still reduce the number further.


```{r, echo=T, messages=F }
# creating a manhattan plot to show which genes would be up regulated and which genes are down regulated in this scenario
rna.dat <- datafile

rna.dat$gene <- as.factor(row.names(rna.dat))
summary(rna.dat)

# separatign bud phase from flower phase so that we can take the mean of each of these
bud.dat <- rna.dat[,c("gene", "Bud_Rep1", "Bud_Rep2", "Bud_Rep3")]
flower.dat <- rna.dat[,c("gene", "Day2_Rep1", "Day2_Rep2", "Day2_Rep3")]

summary(bud.dat)
bud.mean <- data.frame(gene=bud.dat$gene,
                       value = rowMeans(bud.dat[,c("Bud_Rep1", "Bud_Rep2", "Bud_Rep3")]),
                       type = as.factor("bud"))
summary(bud.mean)
bud.mean$log2.value <- ifelse(bud.mean$value == 0, log2(bud.mean$value + 1e-6), log2(bud.mean$value)) # adding a small value to all zero values to avoid infinity error

flower.mean <- data.frame(gene=flower.dat$gene,
                       value = rowMeans(flower.dat[,c("Day2_Rep1", "Day2_Rep2", "Day2_Rep3")]),
                       type = as.factor("flower"))
flower.mean$log2.value <- ifelse(flower.mean$value == 0, log2(flower.mean$value + 1e-6), log2(flower.mean$value)) # adding a small value to all zero values to avoid infinity error
summary(flower.mean)

# stacking together for graphing purposes

rna.mean.stack <- rbind(bud.mean, flower.mean)

# adding color for the up/down regulation of genes
rna.mean.stack$reg <- as.factor(ifelse(rna.mean.stack$gene %in% up.reg$gene, 1,
                             ifelse(rna.mean.stack$gene %in% down.reg$gene,-1,0)) )

summary(rna.mean.stack)

gn.names <- as.factor(matrix(unlist(strsplit(paste(rna.mean.stack$gene), "_")), nrow=nrow(rna.mean.stack), byrow=T)[,2])
rna.mean.stack$pos <- matrix(unlist(strsplit(paste(gn.names), "-")), nrow=nrow(rna.mean.stack), byrow=T)[,1]
rna.mean.stack$pos <- as.numeric(rna.mean.stack$pos)
```


```{r, echo=FALSE, messages=F, fig.height=20, fig.width=15, messages=F, warning=FALSE }
ggplot(data=rna.mean.stack) +  facet_grid(type~.) +
  geom_point(aes(x=pos, y=-log2.value, col=reg)) +
  scale_color_manual(values=c("blue", "grey50", "red"))
```

#### These initial plots were used to identify genes that in the bud stage were completely turned off and then were turned on in the flowering stage, and those genes that were expressed during the bud stage and then completely turned off in the flowering stage. Note: data were -log2 transformed to assist with visualization.

```{r, echo=T, messages=F}
# Pulling out the band of values at the top that are colored red in the bud
# Meaning they were turned off in the bud but were then turned on in the flower

rna.mean.stack.bud <- rna.mean.stack[rna.mean.stack$type=="bud",]
summary(rna.mean.stack.bud)
rna.mean.stack.bud$gn.sig <- as.factor(ifelse(rna.mean.stack.bud$log2.value < -10 & rna.mean.stack.bud$reg == 1,"Y","N"))
summary(rna.mean.stack.bud)

gn.on <- ggplot(rna.mean.stack.bud[rna.mean.stack.bud$gn.sig=="Y",]) +
  geom_hline(aes(, yintercept=1),lwd=1000, col="grey65") +
  geom_vline(aes(xintercept=pos, y=1), col="purple") +
  labs(x="Position", y="", fill="", title="Turned On", caption = "n = 99") +
  theme(axis.line=element_line(color="black"), 
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),  
        panel.background=element_blank(), 
        axis.text.x=element_text(angle=0, color="black", size=16, vjust= 0.5, face="bold"), 
        axis.text.y=element_blank(), 
        strip.text=element_text(face="bold", size=22),
        axis.line.x = element_line(color="black", size = 0.5),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="top",
        legend.key.size = unit(0.75, "cm"),
        legend.text = element_text(size=18),
        legend.title = element_text(size=22),
        legend.key = element_rect(fill = "white")) + 
  guides(fill = guide_colourbar(barwidth = 15, barheight = 3,title="")) +
  theme(axis.title.y= element_text(size=24, face="bold")) +
  theme(axis.title.x= element_text(size=24, face="bold")) +
  scale_x_continuous(limits = c(0,max(rna.mean.stack.bud$pos)),breaks  = seq(0,max(rna.mean.stack.bud$pos), by=10000))

# The center of the distribution is just a flat out mess, but there are blips of blue around the 20 mark that merit investigation
# Pulling out the band of values at the top that are colored blue in the flower meaning they were turned on in teh bud and are not turned off
rna.mean.stack.flower <- rna.mean.stack[rna.mean.stack$type=="flower",]
summary(rna.mean.stack.flower)
rna.mean.stack.flower$gn.sig <- as.factor(ifelse(rna.mean.stack.flower$log2.value < -10 & rna.mean.stack.flower$reg == -1,"Y","N"))

summary(rna.mean.stack.flower)

gn.off <- ggplot(rna.mean.stack.flower[rna.mean.stack.flower$gn.sig=="Y",]) +
  geom_hline(aes(, yintercept=1),lwd=1000, col="grey65") +
  geom_vline(aes(xintercept=pos, y=1), col="forestgreen") +
  labs(x="Position", y="", fill="", title = "Turned Off", caption ="n = 108") +
  theme(axis.line=element_line(color="black"), 
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),  
        panel.background=element_blank(), 
        axis.text.x=element_text(angle=0, color="black", size=16, vjust= 0.5, face="bold"), 
        axis.text.y=element_blank(), 
        strip.text=element_text(face="bold", size=22),
        axis.line.x = element_line(color="black", size = 0.5),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="top",
        legend.key.size = unit(0.75, "cm"),
        legend.text = element_text(size=18),
        legend.title = element_text(size=22),
        legend.key = element_rect(fill = "white")) + 
  guides(fill = guide_colourbar(barwidth = 15, barheight = 3,title="")) +
  theme(axis.title.y= element_text(size=24, face="bold")) +
  theme(axis.title.x= element_text(size=24, face="bold")) +
  scale_x_continuous(limits = c(0,max(rna.mean.stack.bud$pos)),breaks  = seq(0,max(rna.mean.stack.bud$pos), by=10000))

```
#### From the above plots, we can identify the genes that were either switched on or off after flowering present them in a more appealing way.

```{r, echo=FALSE, warning=FALSE, fig.height=8, fig.width=15, messages=F}
library(cowplot)
cowplot::plot_grid(gn.on,gn.off,align = c("v","h"), nrow = 2)
```

#### From the original ca. 30,000 genes we have identified 99 that were significantly up regulated and 108 that were significantly down regulated between bud and flower stages.  At this point, I would rely on the expertise of my colleagues to further refine these candidates for further analysis in a cost-effective manner.





*Code was amended from examples illustrated in 'Getting Genetics Done' website https://www.gettinggeneticsdone.com/2012/09/deseq-vs-edger-comparison.html*

*edgeR documentation can be found here: https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf*